Dynamic Modeling and Parameter Estimation in \proglang{R}

### 2016-09-23 article.tex
Teil eins von Abschnitt 2, Implementation and design principles

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}
\usepackage[]{amsmath}

@@

\subsection{Model formulation}
\subsection{Observation functions}
\subsection{Parameter transformations}
\subsection{Multi-conditional parameter estimation}

@@

\begin{figure}[ht]
	\includegraphics[width = \textwidth]{images/figure1}
	\caption{Output of the prediction function. \textbf{(A)} Prediction of the TCA states. \textbf{(B)} Sensitivities of the three TCA states with respect to the rate parameters.}
	\label{fig:prediction}
\end{figure}

@@

out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Finally, we obtain the simulated data by taking a subset of the prediction. At this point we make use of the "conditions" argument which allows to request a prediction for a vector of conditions, if provided by the prediction function\footnote{Both, \code{x} and \code{g} have been initialized with \code{condition = NULL} which means they are generic prediction functions and yield the same prediction for any of the requested conditions. The concept is explained in Section~\ref{sec:conditions}.}. Data uncertainties $\sigma$ are derived by the Poisson nature of radioactive count experiments, i.e. $\sigma_x = \sqrt{x}$. To avoid division by 0, the minimal $\sigma$-value is set to 1. Random values are added to the predicted values to simulate observation noise. In the end, the \code{data.frame} is converted into a \code{datalist} object.
\begin{CodeChunk}
\begin{CodeInput}
# Simuate data
timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
datasheet <- subset(wide2long(out),
                    time %in% timesD & name %in% names(observables))
datasheet$sigma <- sqrt(datasheet$value + 1)
datasheet$value <- rnorm(nrow(datasheet), datasheet$value, datasheet$sigma)

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the \pkg{parallel} package is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of interations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. \textbf{(A)} Fits have been sorted by increasing least squares value. Four optima were found with almost identical objective value. \textbf{(B)} The parameter values for different optima are shown in different colors }
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B. From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

bestfit <- as.parvec(myframe)
plot((g*x*p)(times, bestfit), data)
\end{CodeInput}
\end{CodeChunk}
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure6}
	\caption{Result of multi-start fitting procedure with two experimental conditions. \textbf{(A)} Model prediction of the best fit and the simulated data are shown in different colors. \textbf{(B)} Fits have been sorted by increasing least squares value. The lowest value clearly separates from the second plateau. \textbf{(C)} Plotting the parameter values for each of the fits reveals that the second plateau consists of two optima. The lowest plateau however corresponds to a unique optimum.}
	\label{fig:twoconditions}
\end{figure}

@@

The code
\begin{CodeChunk}
\begin{CodeInput}
profiles <- profile(obj, bestfit, names(bestfit), limits = c(-5, 5), cores = 4)
plotProfile(profiles)
plotPaths(profiles, whichPar = "s")
\end{CodeInput}
\end{CodeChunk}
gives us the result shown in Figure~\ref{fig:pl}.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure7}
	\caption{Profile likelihood. \textbf{(A)} Profiles of all parameters. Data- and prior contribution to the total objective value are distinguished by line-type. \textbf{(B)} Parameter paths for the scaling parameter \code{s}.}
	\label{fig:pl}
\end{figure}
Computing the profile likelihood, the sum of data contribution, \code{normL2}, and prior contribution, \code{constraintL2}, are optimized under the constraint of a given parameter value for the profiled parameter. In the optimum, data and prior contribution are evaluated separately giving rise to the dashed and dotted lines in Figure~\ref{fig:pl}A. As we had expected, the data contribution to the initial value parameters \code{TCA_cell}, \code{TCA_cana} and the scaling parameter \code{s} is constantly zero. The parameters are structurally non-identifiable.

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}
\usepackage[]{amsmath}
\usepackage{amssymb}

@@

\subsection{Model formulation}

A system of ordinary differential equation $\dot x = f(x, p, u, t)$ can be represented by a \code{character} vector of equations involving symbols for the states $x$ (coinciding with names of the vector), symbols for the parameters $p$, symbols for the forcings $u$ and the keyword \code{time} for $t$. The class provided by \pkg{dMod} for such objects is the \code{eqnvec} class which checks whether the character can be parsed and thus, can be interpreted as an equation.

There are several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tideous to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix [REF]. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step and add further reactions to an already existing \code{eqnlist}. The conversion to the (differential) equations is done by \code{as.eqnvec()} which does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

The paradigm of \pkg{dMod} is to make use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If many parameters are involved and the system consists of many states, the number of equations quickly grows and solving the equations can be considerably accelerated if compiled code is used. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written into the working directory and compiled. The \code{odemodel} object keeps track of the shared objects and is the basis of a prediction function $x(t, p)$ which is generated by the \code{Xs()} command.

\pkg{dMod} seeks to stay close to the mathematical formulation, i.e.~\code{x <- Xs(myodemodel)} will indeed return an \proglang{R} function, an object of class \code{prdfn}, which expects arguments \code{times} and \code{pars} and turns them into a model prediction. However, not only the solution $x(t)$ is returned, the sensitivities $\frac{\partial x}{\partial p}(t)$ are returned, too.

\subsection{Observation functions}

An observation function is a function $g(x, p_{\rm obs}, t)$ that evaluates the solution $x(t)$ together with additional (observation) parameters $p_{\rm obs}$. The function can explicitly depend on time~$t$. Similar to the ODE case, an observation function can be expressed as a character vector with names corresponding to the names of the observables and equations involving symbols for the states, parameters and the keyword \code{time}. An observation function is defined as an \code{eqnvec} and turned into an \proglang{R} function via \code{Y()}.

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied, yielding $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is translated into a \proglang{C} code and compiled by default.

When observation- and prediction functions are concatenated, the result is a prediction function, e.g.~\code{y = g*x} is the \proglang{R} function computing values $y(t)$ from the arguments \code{times} and \code{pars} via evaluation of the ODE and subsequent evaluation of the observation function. Two observation functions can be concatenated, too, yielding an observation function.

\subsection{Parameter transformations}

Parameter transformations are the key element of \pkg{dMod} to formulate different kinds of constraints and allow the combination of several experimental/modeling conditions in one parameter vector.

In principle, a parameter transformation $p = \Phi(\theta)$ is a (differentiable) function connecting \textbf{inner} parameters $p$ with \textbf{outer} parameters $\theta$. The rationale behind the distinction of inner and outer parameters is that the vector $p$ usually desribes those parameters defined in the model equations. The outer parameters $\theta$ refer to a convenient parameterization by which the model parameters are computed.

An \proglang{R} function of class \code{parfn} is produced by the \code{P()} command. Transformations can either be formulated explicitly or implicitly. In the explicit case, the function $p = \Phi(\theta)$ corresponds to an \code{eqnvec} whose names are the names of the inner parameters and entries are equations with symbols for the outer parameters. An implicit transformation has the form $f(p = \Phi(\theta), \theta) = 0$. In this case, $f$ is expressed by an \code{eqnvec} with equations containing symbols for $p$ and $\theta$ and the names of the \code{eqnvec} are the symbols for $p$.

Similar to prediction- and observation functions, parameter functions not only return parameter values but the Jacobian of the transformation, too. Exploiting the chain rule, the derivatives are propagated, allowing to define \code{y = g*x*p} which is a function returning $y(t, \theta)$ and $\frac{\partial y}{\partial \theta}(t)$, i.e.~\code{g*x*p} is a prediction function.

Let \code{g}, \code{x}, \code{p1} and \code{p2} be an observation-, a prediction- and two parameter transformation functions. Then \code{p1*p2} is a parameter transformation function, \code{x*p1} is a prediction function and \code{g*p1} is an observation function.

\subsection{Multi-conditional prediction}

A set of parameter values, forcings and events reflects a certain condition in which we find a dynamic system. By manipulation of the system, single parameters, forcings or events can be changed. It is a typical approach to systems inference to systematically perturb small parts of a dynamic system to reveal information about the underlying process.

The aim of \pkg{dMod} is to allow for "simultaneous" prediction under several conditions and compare these predictions to the corresponding experimental data sets to estimate model parameters. The full information contained in several perturbation experiments is only used if the multi-conditional model prediction is set up in a way that all model parameters not being chaged from condition to condition correspond to exactly one parameter. On the other hand, parameters which are systematically perturbed correspond to different parameters.

Mathematically speaking, we want to construct a parameter transformation
\begin{equation}
	\begin{aligned}
		\Phi:\quad \mathbb R^p & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^q \\
		\theta & \longmapsto  \left(\Phi_1(\theta), \dots, \Phi_n(\theta)\right)
	\end{aligned}
	\label{eq:partrafo}
\end{equation}
where $i = 1, \dots, n$ corresponds to the different conditions. If all parameters are shared throughout all conditions, then $p = q$. If, however, all parameters are distinct, then $p = n\cdot q$. Perturbation experiments correspond to a situation where $p\gtrsim q$.

The \pkg{dMod} package allows to define transformations $\Phi$, see eq.~\eqref{eq:partrafo}, by the \code{"+"} operator: 
\begin{equation*}
\verb!p <- P(eqnvec1, condition = "one") + P(eqnvec2, condition = "two")!. 
\end{equation*}
All symbols from \code{eqnvec1} and \code{eqnvec2} are collected and their union constitutes the symbols of $\theta$. The evaluation \code{p(theta)} returns a list of length two with inner parameters. The \code{"+"} operator can be applied consecutively to add condition by condition.

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)) 
	\end{aligned}
	\label{}
\end{align*}
and the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined and combined by the \code{"+"} operator analogously to \code{p}.

All commands, \code{P()}, \code{Xs()}, \code{Y()}, etc.~can be executed with \code{condition = NULL}. In that case, the corresponding function is generic and, if called for different conditions, the same identical function is evaluated with the condition-specific input.

@@

\begin{figure}[ht]
	\includegraphics[width = \textwidth]{images/figure1}
	\caption{Output of the prediction function. (A) Prediction of the TCA states. (B) Sensitivities of the three TCA states with respect to the rate parameters.}
	\label{fig:prediction}
\end{figure}

@@

out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Finally, we obtain the simulated data by taking a subset of the prediction. At this point we make use of the "conditions" argument which allows to request a prediction for a vector of conditions, if provided by the prediction function\footnote{Both, \code{x} and \code{g} have been initialized with \code{condition = NULL} which means they are generic prediction functions and yield the same prediction for any of the requested conditions. The concept is further illustrated in Section~\ref{sec:conditions}.}. Data uncertainties $\sigma$ are derived by the Poisson nature of radioactive count experiments, i.e. $\sigma_x = \sqrt{x}$. To avoid division by 0, the minimal $\sigma$-value is set to 1. Random values are added to the predicted values to simulate observation noise. In the end, the \code{data.frame} is converted into a \code{datalist} object.
\begin{CodeChunk}
\begin{CodeInput}
# Simuate data
timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
datasheet <- subset(wide2long(out),
                    time %in% timesD & name %in% names(observables))
datasheet$sigma <- sqrt(datasheet$value + 1)
datasheet$value <- rnorm(nrow(datasheet), datasheet$value, datasheet$sigma)

@@

plotValues(myframe, value < 100)
plotPars(myframe, value < 100)
\end{CodeInput}
\end{CodeChunk}
Here, we have searched according to $\vec p_i = \vec p_0 + \Delta\vec p_i$ where $\vec p_0$ is the center, in our case \code{pouter}, $\Delta\vec p_i\sim N(0, \sigma^2)$ is a random parameter vector taken from a normal distribution, in our case $\sigma = 4$, and the index $i$ runs from 1 to \code{fits = 50}. The \code{mclapply()} command from the \pkg{parallel} package is used internally to run fits in parallel, here \code{cores = 4}. The result of \code{mstrust()} is a list of all returned values of \code{trust()}. To extract the final objective value, parameter values, convergence information and the number of interations, \code{as.parframe()} is used. The multi-start approach identifies four local optima, see Figure~\ref{fig:mstrust}, which yield almost the same objective value, Figure~\ref{fig:mstrust}A.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure5}
	\caption{Result of multi-start fitting procedure. (A) Fits have been sorted by increasing least squares value. Four optima were found with almost identical objective value. (B) The parameter values for different optima are shown in different colors }
	\label{fig:mstrust}
\end{figure}
Despite the similar objective value, the optima are not close to each other in parameter space, as being illustrated by Figure~\ref{fig:mstrust}B. From the initial value parameters \code{TCA_cell} and \code{TCA_cana} it seems that they are exchangable. Where \code{TCA_cell} is high, \code{TCA_cana} is low and vice versa. This is not surprising because the observed cellular TCA amount is the sum of both. A new experiment needs to be designed to distinguish one situation from another.

@@

bestfit <- as.parvec(myframe)
plot((g*x*p)(times, bestfit), data)
\end{CodeInput}
\end{CodeChunk}
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure6}
	\caption{Result of multi-start fitting procedure with two experimental conditions. (A) Model prediction of the best fit and the simulated data are shown in different colors. (B) Fits have been sorted by increasing least squares value. The lowest value clearly separates from the second plateau. (C) Plotting the parameter values for each of the fits reveals that the second plateau consists of two optima. The lowest plateau however corresponds to a unique optimum.}
	\label{fig:twoconditions}
\end{figure}

@@

The code
\begin{CodeChunk}
\begin{CodeInput}
profiles <- profile(obj, bestfit, names(bestfit), limits = c(-5, 5), cores = 4)
plotProfile(profiles)
plotPaths(profiles, whichPar = "s")
\end{CodeInput}
\end{CodeChunk}
gives us the result shown in Figure~\ref{fig:pl}.
\begin{figure}[ht]
	\centering
	\includegraphics[width = \textwidth]{images/figure7}
	\caption{Profile likelihood. (A) Profiles of all parameters. Data- and prior contribution to the total objective value are distinguished by line-type. (B) Parameter paths for the scaling parameter \code{s}.}
	\label{fig:pl}
\end{figure}
Computing the profile likelihood, the sum of data contribution, \code{normL2}, and prior contribution, \code{constraintL2}, are optimized under the constraint of a given parameter value for the profiled parameter. In the optimum, data and prior contribution are evaluated separately giving rise to the dashed and dotted lines in Figure~\ref{fig:pl}A. As we had expected, the data contribution to the initial value parameters \code{TCA_cell}, \code{TCA_cana} and the scaling parameter \code{s} is constantly zero. The parameters are structurally non-identifiable.
